### 2021-09-19 plos_latex_template.tex
Rocío: mostly editing and updating references

Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions. 
However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields. %RJ: (Biggs et al. 2018) https://link.springer.com/article/10.1007/s11199-017-0800-9
Conversely, efforts to address systematic discrimination in conferences and make them inclusive spaces can have a meaningful impact in our communities and the lives and careers of their members.

@@

The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, that coordinates with the local organizers of conferences, or even the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive and accessible experience for everyone. 
Here, we share the lessons learned within the past year of organizing this global useR!2021, summarized as 10 simple rules towards an inclusive conference.
% Rocío: Could add a line introducing Fig. in https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing

@@

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. 
Human diversity should be celebrated and respected in every way. Nonetheless, we live in a world with implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems--including conferences--are consciously and unconsciously built.  
%sara: reference for privilege: McIntosh, Peggy. "White privilege: Unpacking the invisible knapsack." (1988).
While no isolated initiative can change reality by itself, building a more diverse and inclusive conference starts by recognizing that these inequalities have excluded systematically whole groups of people from academia and from scientific and professional circles \cite{timperleyHeMoanaPukepuke2020}.

@@

Investing more in effort in the most excluded groups does not mean neglecting the others (see \textbf{Rule \ref{rule_inclusion}} for more on that). 
But it will prevent from falling into the mistake of treating diversity as a checklist.
It will also guide the vision of diversity for your conference--and your strategies to achieve it:
Will a more diverse conference would translate into an even gender distribution in your speakers? 
Would it be representation of all races--especially Black people--among the head organizers, speakers, and attendees? 
Would it be having LGBTQIA+ friendly-spaces or community participation from key geographic regions?
This exercise will also help coming up with indicators that would help the organizing team assess the fulfillment of the diversity objectives along the way. 
%sara: reference for privilege: McIntosh, Peggy. "White privilege: Unpacking the invisible knapsack." (1988).

%batool: There are a few references which define privilege and might be good to add here including: Friedman, S., O’Brien, D., & McDonald, I. (2021). Deflecting Privilege: Class Identity and the Intergenerational Self. Sociology, 55(4), 716–733. https://doi.org/10.1177/0038038520982225
%batool: Another one is: Crevani, L. (2019). Privilege in place: How organisational practices contribute to meshing privilege in place. Scandinavian Journal of Management, 35(2). https://doi.org/10.1016/j.scaman.2018.09.002
%batool: please note that I didn't read both entire papers I mentioned above.

@@

Paying attention to some dimensions of diversity while neglecting others may have unintended, even harmful consequences. Lack of representation, unwelcoming--or overtly aggressive-- environments hinder participation (or future participation) of people who could otherwise become active community members. A negative conference experience can divert career paths, affect lives, and exclude people from some fields. %Rocío: it would be nice to have reference here.
%ast: maybe https://link.springer.com/article/10.1007/s11199-017-0800-9
%Adithi: Assist newbies of the conference navigate through the conference as they might feel overwhelmed attending a conference.
% Rocío: Andrea, could the newbies part go in the networking (box)? I don't see how to make it fit here. 
%i took the networking part to expand communication att.ast we can switch back afterwards

Adopting a code of conduct and creating a team to enforce it are key aspects in creating a safe environment during a conference \cite{favaroYourScienceConference2016}.
The code of conduct is a document meant to keep the community safe and should state clearly: the unacceptable behaviors, the spaces of the conference in which it applies, the consequences for engaging in unacceptable behavior, and the way to report violations \cite{aurora2019}. 
The code of conduct should be displayed prominently in several spaces of the conference to deter people from unacceptable behavior.
An efficient code of conduct acts as a protection for the community because the people who are the target of unacceptable behavior tend to be the ones with less power or privilege.

The code of conduct team should receive training on how to receive reports, respond to incidents, and communicate their responses, and organize accordingly. A diverse code of conduct team will be more understanding of power dynamics and sensitive to discrimination and harassment issues. There are also disciplinary cultures and geographical considerations that may need to be considered when developing and communicating the code of conduct (CoC). We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{aurora2019} as an excellent starting point for the Code of Conduct team work.

%Natalia: the book is by Aurora and Gardiner (2019), I corrected the title and added the reference including the link

@@

The following advice should apply to any kind of team, but are specially relevant when working with people from marginalized groups. First, rethink power relationships inside the group and share power. Be committed to examine power relations inside your organizing team and challenge it during all the process. % Rocío: Could we explain power relationship / share power ideas in more simple terms?
Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer leading positions, and let people from privileged groups step down. 
Build an environment in which every person has their voice and enforce voices from systematically excluded groups.
Offer mentorship and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication (See \textbf{Rule \ref{rule_language}}), or others. You can, for example, break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}). Make work visible, discuss collectively what should be done (and why), divide and account work. Being transparent and sharing the information is key to make work visible and collective. %sara: this relates to make labor visible from DF, maybe we can cite them here too. 
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work.

@@

In-person interaction in conferences is priceless, but only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for global conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. 
% Rocío: we can add our Nature Correspondence letter to the citations later on. 
Online conferences are more inclusive: they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities.
% Rocío: cite Saliba 2020
This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}.
% Rocío: we can add Roosetal. and our Nature Correspondence letter to the citations here too. I would've liked one that talks about increase in participation of tutors and presenters but haven't found any. 
The online format may also make it easier to be inclusive of geographic regions by encompassing several timezones in the week or easily deciding to rotate the favored timezone year to year, without depending on a conference central location. 
% Rocío: main idea of the paragraph: online conferences are more inclusive

Networking and socializing in online conferences may be challenging, mostly because we are used to personal interactions at coffee or lunch breaks, or other in-person social settings. %Rocío: cite Saliba 2020. 
However, social events in in-person conferences can be exclusive (e.g. galas or dinner nights at expensive venues), and they are a common place for code of conduct violations to occur. % Rocío: any reference to back the last statement? The CoC guide?
Organizers should invest time in creating opportunities to meet and bond virtually. 
An online event is a great opportunity to create spaces for people who do not enjoy in person networking. 
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with team work like trivia, offer some events that can be passively enjoyed like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session. Respect people's limits, preferences, and remember that `the usual' does not necessarily work with everyone in your community, and that one single networking activity will never serve the whole community. It is worth trying new and varied activities that might work with subgroups of attendees.
Moreover, virtual communication may make people from minoritized groups feel more included, thus participate more (e.g. CITATION). % RJ: https://link.springer.com/content/pdf/10.1007/s10869-011-9232-7.pdf
% Rocío: Networking paragraph.

@@

If the conference is in person, the venue must be accessible for people with mobility limitations. 
Accommodations such as a quiet space for neurodivergent people should also be provided.
Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
Platforms for conference registration and abstract submission, websites, and chat platforms --if used --should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning for presentations should be available in more than one language if possible.
Do not forget about making your social events accessible (e.g. art shows and movies) and making sure that the materials presented are accessible for all. 
These aspects should be tested well in advance of going live.

@@

Many inclusive practices come with a cost and it is important to add them to the budget when planning the conference. 
Good quality and professional captioning, interpreters or translators, accessible and screen-reader friendly registration tools--and chat platform if the conference is online--, paying your organizing team (see \textbf{Rule \ref{rule_team}}), and code of conduct training, are of the some expenses that you may have to include in your budget. 
Budgets are limited, so it is important to estimate costs in advance and define your priorities.

@@

The ten rules stated here can be adapted depending on the conference format and settings.
When organizing useR!2021, we engaged in most of the practices mentioned here, and learned others along the way, so we share them here as part of our learning process. 
We organized useR! during a global pandemic, and as a team, this was a challenging journey. 
Very early on, we tried to recognize that our resources were limited (both in time of the organizing team and financially). 
We had to let go of certain aspects that were important to many of us. 
From that experience, this is our last message: your conference may not become perfectly inclusive and accessible, but the changes you make will make a difference.
Also, keep working to improve inclusion in the next edition of the conference! to document and balance your work, design surveys, ask for feedback from attendees and organizers. Transfer your constructed knowledge to the next organizing team, and adapt year-to-year to include more dimensions of diversity, to address technical and social changes, and to include insights from the new organizing team. Always remember inclusion is an ambition that organisers can and should continually pursue by allowing room for learning and improvement.
If more conferences and domains apply these rules, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead.
And you will make a change towards healthier, stronger, and more inclusive communities.

@@

\section*{Acknowledgments}
The authors of this piece would like to thank every single member of the organizing team of useR! 2021 [ \url{https://user2021.r-project.org/about/global-team/}] for their valuable contribution to an inclusive conference experience, and the R Foundation for charging us with the organization of useR! 2021 and supporting us through the process.

>>>>

Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions. 
However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.
Conversely, efforts to address systematic discrimination in conferences and make them inclusive spaces can have a meaningful impact in our communities and the lives and careers of their members.

@@

The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, that coordinates with the local organizers of conferences, or even the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive and accessible experience for everyone. 
Here, we share the lessons learned within the past year of organizing this global useR!2021, summarized as 10 simple rules towards an inclusive conference.
% Rocío: Could add a line introducing Fig. in https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing and the whole structure

@@

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. 
Human diversity should be celebrated and respected in every way. Nonetheless, we live in a world with implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems--including conferences--are consciously and unconsciously built.  
While no isolated initiative can change reality by itself, building a more diverse and inclusive conference starts by recognizing that these inequalities have excluded systematically whole groups of people from academia and from scientific and professional circles \cite{timperleyHeMoanaPukepuke2020}.

@@

Investing more in effort in the most excluded groups does not mean neglecting the others (see \textbf{Rule \ref{rule_inclusion}} for more on that). 
But it will prevent from falling into the mistake of treating diversity as a checklist.
It will also guide the vision of diversity for your conference--and your strategies to achieve it:
Will a more diverse conference would translate into an even gender distribution in your speakers? 
Would it be representation of all races--especially Black people--among the head organizers, speakers, and attendees? 
Would it be having LGBTQIA+ friendly-spaces or community participation from key geographic regions?
This exercise will also help coming up with indicators that would help the organizing team assess the fulfillment of the diversity objectives along the way.

%batool: There are a few references which define privilege and might be good to add here including: Friedman, S., O’Brien, D., & McDonald, I. (2021). Deflecting Privilege: Class Identity and the Intergenerational Self. Sociology, 55(4), 716–733. https://doi.org/10.1177/0038038520982225
%batool: Another one is: Crevani, L. (2019). Privilege in place: How organisational practices contribute to meshing privilege in place. Scandinavian Journal of Management, 35(2). https://doi.org/10.1016/j.scaman.2018.09.002
%batool: please note that I didn't read both entire papers I mentioned above.
%sara: reference for privilege: McIntosh, Peggy. "White privilege: Unpacking the invisible knapsack." (1988).
% Rocío: I feel like we don't need to cite any reference about privilege the way it is now. But feel free to disagree.

@@

Paying attention to some dimensions of diversity while neglecting others may have unintended, even harmful consequences. Lack of representation, unwelcoming--or overtly aggressive-- environments hinder participation (or future participation) of people who could otherwise become active community members. A negative conference experience can divert career paths, affect lives, and exclude people from some fields \cite{biggsAcademicConferenceChilly2018}. 
%Adithi: Assist newbies of the conference navigate through the conference as they might feel overwhelmed attending a conference.
% Rocío: Andrea, could the newbies part go in the networking (box)? I don't see how to make it fit here. 
%i took the networking part to expand communication att.ast we can switch back afterwards

Adopting a code of conduct and creating a team to enforce it are key aspects in creating a safe environment during a conference \cite{favaroYourScienceConference2016}.
The code of conduct is a document meant to keep the community safe and should state clearly: the unacceptable behaviors, the spaces of the conference in which it applies, the consequences for engaging in unacceptable behavior, and the way to report violations \cite{auroraHowRespondCode2019}. 
The code of conduct should be displayed prominently in several spaces of the conference to deter people from unacceptable behavior.
An efficient code of conduct acts as a protection for the community because the people who are the target of unacceptable behavior tend to be the ones with less power or privilege.

The code of conduct team should receive training on how to receive reports, respond to incidents, and communicate their responses, and organize accordingly. A diverse code of conduct team will be more understanding of power dynamics and sensitive to discrimination and harassment issues. There are also disciplinary cultures and geographical considerations that may need to be considered when developing and communicating the code of conduct (CoC). We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for the Code of Conduct team work.

@@

The following advice should apply to any kind of team, but are specially relevant when working with people from marginalized groups. First, rethink power relationships inside the group and share power. Be committed to examine power relations inside your organizing team and challenge it during all the process. % Rocío: Could we explain power relationship / share power ideas in more simple terms?
Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer leading positions, and let people from privileged groups step down. 
Build an environment in which every person has their voice and enforce voices from systematically excluded groups.
Offer mentorship and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication (See \textbf{Rule \ref{rule_language}}), or others. You can, for example, break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}). Make work visible, discuss collectively what should be done (and why), divide and account work. Being transparent and sharing the information is key to make work visible and collective. %sara: this relates to make labor visible from DF, maybe we can cite them here too. %Rocío: sure.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work.

@@

In-person interaction in conferences is priceless, but only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for global conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. 
% Rocío: we can add our Nature Correspondence letter to the citations later on. 
Online conferences are more inclusive: they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities \cite{salibaGettingGripsOnline2020}.
This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021, roosOnlineConferencesNew2020}.
% Rocío: we can add our Nature Correspondence letter to the citations here too. I would've liked one that talks about increase in participation of tutors and presenters but haven't found any. 
The online format may also make it easier to be inclusive of geographic regions by encompassing several timezones in the week or easily deciding to rotate the favored timezone year to year, without depending on a conference central location. 
% Rocío: main idea of the paragraph: online conferences are more inclusive

Networking and socializing in online conferences may be challenging, mostly because we are used to personal interactions at coffee or lunch breaks, or other in-person social settings \cite{salibaGettingGripsOnline2020}. 
However, social events in in-person conferences can be exclusive (e.g. galas or dinner nights at expensive venues), and they are a common place for code of conduct violations to occur. % Rocío: any reference to back the last statement? The CoC guide?
Organizers should invest time in creating opportunities to meet and bond virtually. 
An online event is a great opportunity to create spaces for people who do not enjoy in person networking. 
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with team work like trivia, offer some events that can be passively enjoyed like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session. Respect people's limits, preferences, and remember that `the usual' does not necessarily work with everyone in your community, and that one single networking activity will never serve the whole community. It is worth trying new and varied activities that might work with subgroups of attendees.
Moreover, virtual communication may make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012}).
% Rocío: Networking paragraph.

@@

If the conference is in person, the venue must be accessible for people with mobility limitations. 
Accommodations such as a quiet space for neurodivergent people should also be provided.
Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
Platforms for conference registration and abstract submission, websites, and chat platforms--if used--should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning for presentations should be available in more than one language if possible.
Do not forget about making your social events accessible (e.g. art shows and movies) and making sure that the materials presented are accessible for all. 
These aspects should be tested well in advance of going live.

@@

Many inclusive practices come with a cost and it is important to add them to the budget when planning the conference. 
Good quality and professional captioning, interpreters or translators, accessible and screen-reader friendly registration tools--and chat platform if the conference is online--, paying your organizing team (see \textbf{Rule \ref{rule_organizing_team}}), and code of conduct training, are of the some expenses that you may have to include in your budget. 
Budgets are limited, so it is important to estimate costs in advance and define your priorities.

@@

The ten rules stated here can be adapted depending on the conference format and settings.
When organizing useR!2021, we engaged in most of the practices mentioned here, and learned others along the way, so we share them here as part of our learning process. 
We organized useR! during a global pandemic, and as a team, this was a challenging journey. 
Very early on, we tried to recognize that our resources were limited (both in time of the organizing team and financially). 
We had to let go of certain aspects that were important to many of us. 
From that experience, this is our last message: your conference may not become perfectly inclusive and accessible, but the changes you make will make a difference.
Also, keep working to improve inclusion in the next edition of the conference! to document and balance your work, design surveys, ask for feedback from attendees and organizers. Transfer your constructed knowledge to the next organizing team, and adapt year-to-year to include more dimensions of diversity, to address technical and social changes, and to include insights from the new organizing team. Always remember inclusion is an ambition that organizers can and should continually pursue by allowing room for learning and improvement.
If more conferences and domains apply these rules, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead.
And you will make a change towards healthier, stronger, and more inclusive communities.

@@

\section*{Acknowledgments}
The authors of this piece would like to thank every single member of the organizing team of useR!2021 [ \url{https://user2021.r-project.org/about/global-team/}] for their valuable contribution to an inclusive conference experience, and the R Foundation for charging us with the organization of useR!2021 and supporting us through the process.
